# Cosinor Analysis {#cosinor}

The issue with time series analysis is that the data is by its nature _circular_ and thus cannot be easily be analyzed through traditional, linear methods. The following is the development/expansion of the cosinor model to help study circadian rhythms (\@ref(circadian)) using `R`. 

## Overview

The `card` package was developed to help tackle this problem. 

```{r}
# Library
library(card)
library(tidyverse)

# Dataset
data("twins")

# Example of data
ggplot(twins, aes(x = hour, y = rDYX)) +
	geom_smooth(method = "gam", se = TRUE) 
```

Using the `cosinor()` function, the characteristics of the circadian pattern can be retrieved.

```{r}
# Cosinor model
m <- cosinor(rDYX ~ hour, twins, tau = 24)
summary(m)
```

The statistical principles behind this method allow for different methods to model, diagnose, and interpret findings. [@Refinetti2007; @Cornelissen2014]

- single component cosinor
- multiple component cosinor
- population cosinor
- confidence intervals (ellipse method)
- zero amplitude test
- lack-of-fit testing

The example here use the dataset `twins` which contains a continuous ECG signal, called `DYX`, collected at hourly time points.

## Single component cosinor

The single component cosinor method is modeled as:

$$Y(t) = M + A cos(\frac{2 \pi t}{\tau} + \phi) + \epsilon$$

Where:

$$
\begin{aligned}
M &= MESOR\ (midline\ estimating\ statistic\ of\ rhythm) \\
t &= time in hours \\
\epsilon &= error \\
\phi &= acrophase \\
\tau &= tau\ (period) \\
\end{aligned}
$$

To model this function, it must be transformed linearly to assess the coefficients.

$$Y(t) = M + \beta x_{t} + \gamma z_{t} + \epsilon_{t}$$

The new coefficients and parameters are defined as: 

$$
\begin{aligned}
\beta &= A cos(\phi) \\
\gamma &= -A sin(\phi) \\
x_{t} &= cos(\frac{2 \pi t}{\tau}) \\
z_{t} &= sin(\frac{2 \pi t}{\tau}) \\
\end{aligned}
$$

In the `twins` data, the time value $t$ is measured in hours. As this is 24-hour data, the assumption is that $\tau = 24$. 

```{r}
df <- subset(twins, patid == 60) # Single individual cosinor
y <- df$rDYX
t <- df$hour
n <- length(t) # Number of observations
period <- 24

# Transformed variables
x <- cos(( 2 * pi * t) / period)
z <- sin(( 2 * pi * t) / period)
```

To generate the coefficients in `R` requires sovling a matrix of normal/linear equations. 

```{r}
# Matrices
ymat <- as.matrix(cbind(y = c(sum(y), sum(y * x), sum(y * z))))
mcol <- c(n, sum(x), sum(z)) # Mesor column
bcol <- c(sum(x), sum(x^2), sum(x * z)) # Beta column
gcol <- c(sum(z), sum(x * z), sum(z^2)) # Gamma column
xmat <- as.matrix(cbind(m = mcol, b = bcol, g = gcol))

# Solution
coefs <- solve(t(xmat) %*% xmat, tol = 1e-21) %*% (t(xmat) %*% ymat)
mesor <- coefs[1] # mesor
beta <- coefs[2] # beta
gamma <- coefs[3] # gamma
```

For a single cosinor, as in, the analysis of the values from a single individual over 1 period, the values for the _Amplitude_ ($A$) and _Acrophase_ ($\phi$) can be calculated.

$$
\begin{aligned}
A &= \sqrt{(\beta^2 + \gamma^2)}
\phi &= k \pi + g \times arctan(\frac{\gamma}{\beta})
\end{aligned}
$$

Because the values of $\gamma$ and $\beta$ represent trigonemtric values, the position or quadrant of the circle changes the value of $\phi$.

| $\beta$ | $\gamma$ | k | g |
| --- | --- | --- | --- |
| + | + | 0 | -1 |
| + | - | $-2 \pi$ | +1 |
| - | + | $- \pi$ | +1 |
| - | - | $- \pi$$ | -1 |

These calculations were made with the cosinor model seen above. 

```{r}
# Amplitude
amp <- sqrt(beta^2 + gamma^2)

# Acrophase (phi) must be in correct quadrant
sb <- sign(beta)
sg <- sign(gamma)
theta <- atan(abs(gamma / beta))

if ((sb == 1 | sb == 0) & sg == 1) {
  phi <- -theta
} else if (sb == -1 & (sg == 1 | sg == 0)) {
  phi <- theta - pi
} else if ((sb == -1 | sb == 0) & sg == -1) {
  phi <- -theta - pi
} else if (sb == 1 & (sg == -1 | sg == 0)) {
  phi <- theta - (2 * pi)
}

cat(paste0("Amplitude = ", round(amp, 3)))
cat(paste0("Acrophase = ", round(phi, 3)))
```

## Population-mean cosinor

Based on the work by Cornelissen et al 2014 [@Cornelissen2014], the population mean cosinor can be estimated by applying the single or multiple component cosinor to each individual. 

$$\{\hat{u} = \hat{M}_{i} + \hat\beta_{i} + \hat\gamma_{i} + ... \}$$

Where $i = 1, 2, ..., k$ for each individual contribution to the population cosinor metrics.

Each parameter can then be "averaged" to estimate the population parameters. This allows extension from a single individual to populations, particularly research studies with cohorts of patients. 

The $A$ and $\phi$ however are calculated using the previous equations but through the $\mu_{\beta}$ and $\mu_{\gamma}$ values. The _MESOR_ can be calculated simply by measure the mean value from each sample ($MESOR_{population} = MESOR_{1} + ... + MESOR_{k}$).

```{r}
# Parameters for population mean cosinor, using best datasets
df <- twins %>%
	filter(med_beta_blockers != 1) %>%
	select(c("rDYX", "hour", "patid"))
names(df) <- c("y", "t", "pop")
highCounts <- df %>%
	group_by(pop) %>%
	tally() %>%
	filter(n > 20)

# Subset for full data
df <- subset(df, pop %in% highCounts$pop)

# Number of individuals
k <- length(unique(df$pop))

# Individual cosinor models are implemented for each individual
kCosinors <- 
	with(
		df,
		by(df, pop, function(x) {
			cosinor(y ~ t, data = x, tau = 24)
		})
	)

# The coefficients have to be extracted and summarized
tbl <- sapply(kCosinors, stats::coef)
coef_names <- c("mesor", "amp", "phi", "beta", "gamma")
rownames(tbl) <- coef_names
xmat <- t(tbl)

# Get mean for each parameter (mesor, beta, gamma), ignoring averaged amp/phi
coefs <- apply(xmat, MARGIN = 2, function(x) {
  sum(x) / k
})

mesor <- unname(coefs["mesor"])
beta <- unname(coefs["beta"])
gamma <- unname(coefs["gamma"])

# Get amplitude
amp <- sqrt(beta^2 + gamma^2)

# Acrophase (phi) must be in correct quadrant
sb <- sign(beta)
sg <- sign(gamma)
theta <- atan(abs(gamma / beta))

if ((sb == 1 | sb == 0) & sg == 1) {
  phi <- -theta
} else if (sb == -1 & (sg == 1 | sg == 0)) {
  phi <- theta - pi
} else if ((sb == -1 | sb == 0) & sg == -1) {
  phi <- -theta - pi
} else if (sb == 1 & (sg == -1 | sg == 0)) {
  phi <- theta - (2 * pi)
}

# Update coefficients
coefs["amp"] <- amp
coefs["phi"] <- phi

# Updated coefficients
names(coefs) <- coef_names
print(coefs)
```

### Confidence Intervals for Population Cosinor

The confidence intervals for a population are more complicated to generate, and several approaches are documented in the literature.

#### ELlipsoid Approach

The values, including standard deviation and standard error for the _MESOR_ are calculated using standard statistics along a t-distribution, with degree of freedom based on number of observations. In this case, $\alpha = 0.05$.

```{r}
# Standard error for mesor
kcoefs <- data.frame(xmat)
se <- sd(kcoefs$mesor) / sqrt(k - 1)
cat(round(se, 3))
```

The statistical parameters around the $A$ and $\phi$ are more complex, as they are joined together, and represent a joint confidence region of the substitute parameters $\beta$ and $\gamma$. 

The first step is the calculation of the variance and covariance of $\beta$ and $\gamma$. This can be used to generated teh standard deviation of these variables.

$$
\begin{aligned}
\sigma_{\beta \gamma} &= \sqrt{COV_{\beta \gamma}} \\
\sigma_{\beta} &= \sqrt{VAR_{\beta}} \\
\sigma_{\gamma} &= \sqrt{VAR_{\gamma}} \\
\end{aligned}
$$

```{r}
sbg <- sqrt(cov(kcoefs$beta, kcoefs$gamma))
sb <- sqrt(var(kcoefs$beta))
sg <- sqrt(var(kcoefs$gamma))
```

The next step is the creation of a confidence ellipse for a given confidence interval. This ellipse is defined by all points $(\beta*, \gamma*)$ that satisfy the elliptical equation.

$$
\frac{(\beta - \beta*)^2}{\sigma^2_{\beta}} -
\frac{2r(\beta - \beta*)(\gamma - \gamma*)}{\sigma_{\beta} \sigma_{\gamma}} +
\frac{(\gamma - \gamma*)}{\sigma^2_{\gamma}} = 
\frac{2(1 - r^2)(k - 1)F_{1 - \alpha}}{k(k - 2)}
$$

This can be reorganized/reorderd by solving for a single parameter first, such as $\beta*$, which will lead to two potential values.

$$
\beta* = 
\frac{
	\beta \sigma_{\gamma} - r \sigma_{beta} \gamma + r \sigma_{beta} \gamma* \pm
	\sqrt{(r^2 - 1)[(\gamma* - \gamma)^2 - \frac{2(k-1)}{k(k-2)}F_{1-\alpha}\sigma^2_{\gamma}]}
}
{\sigma_{\gamma}}
$$

... where $r = \frac{\sigma_{\beta \gamma}}{\sigma_{\beta} \sigma_{\gamma}}$

This is calculated using the above equation for a potential sequence of values of $\beta*$ and $\gamma*$. THe constant values are already known to us, including the $\beta$ and $\gamma$ variables.

```{r}
# Variance/covariance and initial values were found above
# Define new constants
alpha <- 0.05
r <- sbg / (sb * sg)
fstat <- qf(1 - alpha, 2, k - 2)

# Sequence values
gseq <- seq(from = -abs(gamma*3), to = abs(gamma*3), length.out = 100)
bpos <- ((beta * sg) - (r * sb * gamma) + (r * sb * gseq) + (sb * sqrt(as.complex((r^2 - 1) * ((gseq - gamma)^2 - ((2 * (k - 1)) / (k * (k - 2)) * fstat * sg^2)))))) / sg
bneg <- ((beta * sg) - (r * sb * gamma) + (r * sb * gseq) - (sb * sqrt(as.complex((r^2 - 1) * ((gseq - gamma)^2 - ((2 * (k - 1)) / (k * (k - 2)) * fstat * sg^2)))))) / sg

# Restrict to only real numbers (not complex/imaginary)
index <- Im(bpos) == 0 | Im(bpos) == Im(bneg) # values are zero in both are REAL numbers
gseq <- Re(gseq[index])
bpos <- Re(bpos[index])
bneg <- Re(bneg[index])


# Plot out ellipse 
ggplot() + 
	# Original values
	geom_point(aes(x = gamma, y = beta), data = kcoefs, alpha = 0.2) + 
	# Potential ellipse versus hyperbola
	geom_point(aes(x = gseq, y = bpos), col = "red", size = 0.5) + 
	geom_point(aes(x = gseq, y = bneg), col = "blue", size = 0.5) + 
	# Predicted segment
	geom_segment(aes(x = 0, y = 0, xend = -amp*sin(phi), yend = amp*cos(phi)), size = 1.5) + 
	# Axes
	geom_vline(xintercept = 0) + 
	geom_hline(yintercept = 0) +
	xlim(-abs(gamma)*5, abs(gamma)*5) + 
	ylim(-abs(beta)*5, abs(beta)*5)


# Using {car}
border <- car::dataEllipse(cbind(kcoefs$gamma, kcoefs$beta), levels = 0.95) %>%
	as_tibble()

ggplot() +
	geom_point(aes(x = x, y = y), data = border, col = "red") + 
	geom_point(aes(x = gamma, y = beta), data = kcoefs, alpha = 0.5) + 
	geom_vline(xintercept = 0) + 
	geom_hline(yintercept = 0) 
```

#### Sampling Matrix Approach

An approach, according to _Bingham et al 1982_, is to use the sampling matrix, generated from the following formulas and calculated below. The key formulas for the population cosinor confidence intervals are:

$$
\begin{aligned}

MESOR &\pm \frac{t_{1 - \alpha/2} \sigma_{MESOR}}{\sqrt{k}} \\

A &\pm t_{1 - \alpha/2} \sqrt{s_{22}} \\

\phi &+ arctan(\frac{s_{23} t^2_{1 - alpha/2} \pm t_{1 - \alpha/2} \sqrt{s_{33}} \sqrt{A^2 - \frac{(s_{22} s_{33} - s^2_{23})t^2_1{1 - \alpha/2}}{s_{33}}}}{A^2 - s_{22} t^2_{1 - \alpha/2}})

\end{aligned}
$$

Where the matrix variables are shown below:

$$
\begin{aligned}
s_{22} &= \frac{\sigma^2_\beta \beta^2 + 2 \sigma_{\beta \gamma} \beta \gamma + \sigma^2_{\gamma} \gamma^2}{k A^2} \\

s_{23} &= \frac{-(\sigma^2_{\beta} - \sigma^2_{\gamma})(\beta \gamma) + \sigma_{\beta \gamma} (\beta^2 - \gamma^2)}{k A^2} \\

s_{33} &= \frac{\sigma^2_{\beta} - 2 \sigma_{\beta \gamma} \beta \gamma + \sigma^2_{\gamma} \beta^2}{k A^2}
	
\end{aligned}
$$

Thus, we can use these equations to calculate the confidence intervals.

```{r}
# Stats
alpha <- 0.05
tdist <- qt(1 - alpha/2, k - 1)

# Matrix variables
s22 <- ((sb^2 * beta^2) + (2 * sbg * beta * gamma) + (sg^2 * gamma^2)) / (k * amp^2)
s23 <- (-1 * (sb^2 - sg^2) * (beta * gamma) + sbg * (beta^2 - gamma^2)) / (k * amp^2)
s33 <- ((sb^2 * gamma^2) - (2 * sbg * beta * gamma) + (sg^2 * beta^2)) / (k * amp^2)
```

#### Approach by Fernandez [@Fernandez2004]

The population aproach can also be predicted through an alternative, perhaps more intuitive way. If normality is assumed, the estimated parameters can be generated from the individual parameters, similar to the _MESOR_, in a single population, and allows for simple statistical testing between populations. [@Fernandez2004]

```{r}
# Stats
kcoefs <- data.frame(xmat)
alpha <- 0.05
tdist <- qt(1 - alpha/2, k - 1)

# Plot
ggplot() +
	geom_segment(
		aes(
			x = gamma - (tdist * sd(kcoefs$gamma) / sqrt(k)),
			xend = gamma + (tdist * sd(kcoefs$gamma) / sqrt(k)),
			y = 0,
			yend = 0
		),
		col = "cornflowerblue",
		size = 2
	) + 
	geom_segment(
		aes(
			y = beta - (tdist * sd(kcoefs$beta) / sqrt(k)),
			yend = beta + (tdist * sd(kcoefs$beta) / sqrt(k)),
			x = 0,
			xend = 0
		),
		col = "indianred",
		size = 2
	) +
	geom_rect(
		aes(
			ymin = beta - (tdist * sd(kcoefs$beta) / sqrt(k)),
			ymax = beta + (tdist * sd(kcoefs$beta) / sqrt(k)),
			xmin = 0,
			xmax = gamma + (tdist * sd(kcoefs$gamma) / sqrt(k))
		),
		fill = "indianred",
		alpha = 0.5
	) + 
	geom_rect( 
		aes(
			ymin = 0,
			ymax = beta - (tdist * sd(kcoefs$beta) / sqrt(k)),
			xmin = gamma - (tdist * sd(kcoefs$gamma) / sqrt(k)),
			xmax = gamma + (tdist * sd(kcoefs$gamma) / sqrt(k))
		),
		fill = "cornflowerblue",
		alpha = 0.5
	) + 
	geom_point(aes(x = gamma, y = beta), size = 2) + 
	geom_segment(
		aes(
			x = 0, 
			y = 0, 
			xend = -amp*sin(phi), 
			yend = amp*cos(phi)
		), 
		size = 1.2
	) + 
	geom_vline(xintercept = 0) + 
	geom_hline(yintercept = 0) 
	
# Values possible for amplitude
is.logical(round(kcoefs$amp * cos(kcoefs$phi), 3) == round(kcoefs$beta, 3))

# Values for phi
is.logical(round(kcoefs$amp * -1 * sin(kcoefs$phi), 3) == round(kcoefs$gamma, 3))

sd(kcoefs$beta)
sd(kcoefs$gamma)
```

However, this method is more complicated when multiple components are included.

## References
