# Cosinor Analysis {#cosinor}

The issue with time series analysis is that the data is by its nature _circular_ and thus cannot be easily be analyzed through traditional, linear methods. The following is the development/expansion of the cosinor model to help study the circadian rhythm \@ref(circadian). 

## Overview

The `card` package was developed to help tackle this problem. 

```{r}
# Library
library(card)
library(ggplot2)

# Dataset
data("twins")

# Example of data
ggplot(twins, aes(x = hour, y = rDYX)) +
	geom_smooth(method = "gam", se = TRUE) 
```

Using the `cosinor()` function, the characteristics of the circadian pattern can be retrieved.

```{r}
# Cosinor model
m <- cosinor(rDYX ~ hour, twins)
summary(m)
```

The statistical principles behind this method allow for different methods to model, diagnose, and interpret findings. [@Refinetti2007; @Cornelissen2014]

- single component cosinor
- multiple component cosinor
- population cosinor
- zero amplitude test
- lack-of-fit testing
- confidence intervals (ellipse method)

## Single component cosinor

The single component cosinor method is modeled as:

$$Y(t) = M + A cos(2 \pi t / period) + \epsilon$$
