--- 
title: "Thoughts"
author: "Anish Shah"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: "asshah4/book"
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# Prerequisites

This is not an organize series of chapters yet. This is likely a _living book_ that will change over time. There will be likely several sections

- __technical__: involving Rmarkdown, Github, Vim, MATLAB
- __physiology__: medically related topics
- __ecg__: concepts taken from signal processing efforts, including morphology and heart rate variability analyses

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

## This project is built with several libraries from CRAN

- bookdown
- pkgdown
